### 2021-05-10 content/content_with_teaser.tex
init

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v2.02, Jan 25, 2017

\title[Seminar Topic]%
      {Seminar Topic about Graph Layouting for Visualizing Graph Structured Data}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
\author[M.Rosier \& K.\"ohlein]
{\parbox{\textwidth}{\centering Marcel Rosier$^1$% 
\\ Seminar: Data Visualization %
\\ marcel.rosier@tum.de% 
\\ Supervisor: Kevin H\"ohlein%
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
        }
        \\
% For Computer Graphics Forum: Please use the abbreviation of your first name.
{\parbox{\textwidth}{\centering $^1$ Technische Universit\"at M\"unchen
%        $^2$ Another Department to illustrate the use in papers from authors
%             with different affiliations
       }
}
}
% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{36}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}

\teaser{
 \includegraphics[width=\linewidth]{eg_new}
 \centering
  \caption{New EG Logo}
\label{fig:teaser}
}

\maketitle
%-------------------------------------------------------------------------
\begin{abstract}
   The ABSTRACT is to be in fully-justified italicized text, 
   between two horizontal lines,
   in one-column format, 
   below the author and affiliation information. 
   Use the word ``Abstract'' as the title, in 9-point Times, boldface type, 
   left-aligned to the text, initially capitalized. 
   The abstract is to be in 9-point, single-spaced type.
   The abstract may be up to 3 inches (7.62 cm) long. \\
   Leave one blank line after the abstract, 
   then add the subject categories according to the ACM Classification Index 
%-------------------------------------------------------------------------
%  ACM CCS 1998
%  (see http://www.acm.org/about/class/1998)
% \begin{classification} % according to http:http://www.acm.org/about/class/1998
% \CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
% \end{classification}
%-------------------------------------------------------------------------

\end{abstract}  
%-------------------------------------------------------------------------
\section{Introduction}

Please follow the steps outlined in this document very carefully when
submitting your manuscript to Eurographics.

You may as well use the \LaTeX\ source as a template to typeset your own
paper. In this case we encourage you to also read the \LaTeX\ comments
embedded in the document.

%-------------------------------------------------------------------------
\section{Instructions}

Please read the following carefully.

%-------------------------------------------------------------------------
\subsection{Language}

All manuscripts must be in English.

%-------------------------------------------------------------------------
\subsection{Margins and page numbering}

All printed material, including text, illustrations, and charts,
must be kept within a print area %6.31 inches (16.03 cm) 
7 inches (17.7 cm) wide by
%9.10 inches (23.13 cm) 
9.44 inches (24 cm) high. Do not write or print anything
outside the print area. Number your pages on odd sites right
above, on even sites left above, no page number on the first site.
Do not use page numbering within the final version of your paper.

%------------------------------------------------------------------------
\subsection{Formatting your paper}

% All text with the exception of the abstract must be in a two-column format.
% The total allowable width of the text area -- including header and footer
% lines -- is 161\,mm (6.34 inch) wide by 231\,mm (9.10 inch) high.
% 
% Columns are to be 76\,mm (3.0 inch) wide, with a 8\,mm (0.315 inch) space 
% between them.

All text with the exception of the abstract must be in a two-column format.
The total allowable width of the text area -- including header and footer
lines -- is 177\,mm (7 inch) wide by 245\,mm (9.64 inch) high.

Columns are to be 84\,mm (3.3 inch) wide, with a 8\,mm (0.315 inch) space 
between them.

The space between the header line and the first line of the text body and
between the last line of the text body and the footer line is 5\,mm
(0.196 inch) each.

%-------------------------------------------------------------------------
\subsection{Type-style and fonts}

Wherever Times is specified, Times Roman may also be used. If
neither is available on your word processor, please use the font
closest in appearance to Times that you have access to. Only
Type-1 fonts will be accepted.

MAIN TITLE. The title should be in Times 17-point, boldface type and
centered. Capitalize the first letter of nouns, pronouns, verbs, adjectives,
and adverbs; do not capitalize articles, coordinate conjunctions, or
prepositions (unless the title begins with such a word). Leave two blank
lines after the title.

AUTHOR NAME(s) and AFFILIATION(s) are to be centered beneath the title and
printed in Times 9-point, non-boldface type. This information is to be
followed by two blank lines.

The ABSTRACT ist to be in a one-column format. The MAIN TEXT is to be in a
two-column format.

MAIN TEXT. Type main text in 9-point Times, single-spaced. Do \emph{not} use
double-spacing. All paragraphs should be indented 1 em (the length of the
dash in the actual font). Make sure your text is fully justified -- that is,
flush left and flush right. Please do not place any additional blank lines
between paragraphs. Figure and table captions should be 9-point Times
boldface type as in Figure~\ref{fig:firstExample}.

\noindent Long captions should be set as in Figure~\ref{fig:ex1} or
Figure~\ref{fig:ex3}.

\begin{figure}[htb]
   % an empty figure just consisting of the caption lines
   \caption{\label{fig:ex1}
     'Empty' figure only to serve as an example of long caption requiring 
     more than one line. It is not typed centered but aligned on both sides.}
\end{figure}

\noindent
Figures which need the full textwidth can be typeset as Figure~\ref{fig:ex3}.

\noindent Callouts should be 9-point Times, non-boldface type. Initially
capitalize only the first word of section titles and first-, second-, and
third-order headings.

FIRST-ORDER HEADINGS. (For example, \textbf{1. Introduction}) should be Times
9-point boldface, initially capitalized, flush left, with one blank line
before, and one blank line after.

SECOND-ORDER HEADINGS. (For example, \textbf{2.1. Language}) should be Times
9-point boldface, initially capitalized, flush left, with one blank line
before, and one after. If you require a third-order heading (we discourage
it), use 9-point Times, boldface, initially capitalized, flush left, preceded
by one blank line, followed by a period and your text on the same line.

The headline \emph{(authors / title)} must be shortened if it uses the full 
two column width of the main text.
There must be enough space for the page numbers. Please use ``et al.'' if 
there are more than three authors and specify a shortened version for your title.
%-------------------------------------------------------------------------
\subsection{Footnotes}

Please do \emph{not} use footnotes at all!

%-------------------------------------------------------------------------
\subsection{References}

List all bibliographical references in 9-point Times, single-spaced, at the
end of your paper in alphabetical order. When referenced in the text, enclose
the citation index in square brackets, for example~\cite{herman2000graph}. Where
appropriate, include the name(s) of editors of referenced books.

For your references please use the following algorithm:
\begin{itemize} 
\item \textbf{one} author: first 3 chars plus year -- 
      e.g.\ \cite{Herman2000} , \cite{VonLandesberger2011} \cite{Goyal2018} \cite{Lhuillier2017} \cite{Vehlow2015} \cite{Gibson2013}
\item \textbf{two}, \textbf{three} or \textbf{four} authors: first char
      of each family name plus year --  e.g.\ \cite{Fellner-Helmberg93} 
      or \cite{Kobbelt97-USHDR} or \cite{Lafortune97-NARF}
\item \textbf{more than 4} authors: first char of family name from 
      first 3 authors followed by a '*' followed by the year -- 
      e.g.\ \cite{Buhmann:1998:DCQ} or \cite{FolDamFeiHug.etal93} 
\end{itemize}

For BibTeX users a style file \ \texttt{eg-alpha.bst} \ is available which
uses the above algorithm.

%-------------------------------------------------------------------------
\subsection{Illustrations, graphs, and photographs}

All graphics should be centered.

%%%
%%% Figure 1
%%%
\begin{figure}[htb]
  \centering
  % the following command controls the width of the embedded PS file
  % (relative to the width of the current column)
  \includegraphics[width=.8\linewidth]{sampleFig}
  % replacing the above command with the one below will explicitly set
  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
  % It will also prevent LaTeX from reading the PS file to determine
  % the bounding box (i.e., it will speed up the compilation process)
  % \includegraphics[width=.95\linewidth, bb=39 696 126 756]{sampleFig}
  %
  \parbox[t]{.9\columnwidth}{\relax
           For all figures please keep in mind that you \textbf{must not}
           use images with transparent background! 
           }
  %
  \caption{\label{fig:firstExample}
           Here is a sample figure.}
\end{figure}

If your paper includes images, it is very important that they are of
sufficient resolution to be faithfully reproduced.

To determine the optimum size (width and height) of an image, measure
the image's size as it appears in your document (in millimeters), and
then multiply those two values by 12. The resulting values are the
optimum $x$ and $y$ resolution, in pixels, of the image. Image quality
will suffer if these guidelines are not followed.

Example 1: 
%
An image measures 50\,mm by 75\,mm when placed in a document. This
image should have a resolution of no less than 600 pixels by 900
pixels in order to be reproduced faithfully.

Example 2: 
%
Capturing a screenshot of your entire $1024 \times 768$ pixel display
monitor may be useful in illustrating a concept from your research. In
order to be reproduced faithfully, that $1024 \times 768$ image should
be no larger than 85 mm by 64 mm (approximately) when placed in your
document.

%-------------------------------------------------------------------------
\subsection{Color}

\textbf{Please observe:} as of 2003 publications in the proceedings of the
Eurographics Conference can use color images throughout the paper. No
separate color tables are necessary.

However, workshop proceedings might have different agreements! 
Figure~\ref{fig:ex3} is an example for creating color plates.

%------------------------------------------------------------------------
\subsection{Embedding of Hyperlinks / Typesetting of URLs}

Due to the use of the package \texttt{hyperref} the original behavior
of the command $\backslash$\texttt{url} from the package \texttt{url}
is not available. To circumvent this problem we either recommend to
use the command $\backslash$\texttt{httpAddr} from the 
included package \texttt{egweblnk} (see below) or to replace the
command $\backslash$\texttt{url} by the command $\backslash$\texttt{webLink} 
-- e.g. in cases where $\backslash$\texttt{url} has been used
widely in BibTeX-References. In the latter case we suggest to run
BibTeX as usual and then replace all occurences of $\backslash$\texttt{url}  by
$\backslash$\texttt{webLink}

\noindent
The provided commands for hyperlinks, in a nutshell, are:

\begin{description} \itemsep 1ex
\item [\webLinkFont $\backslash$httpAddr \{URL without leading 'http:'\}]
      \mbox{}\\
      e.g. \  \httpAddr{//diglib.eg.org/EG/DL/WS}

\item[\webLinkFont $\backslash$ftpAddr \{URL without leading 'ftp:'\}]
      \mbox{}\\
      e.g. \  \ftpAddr{//www.eg.org/EG/DL/ftpupload}

\item[\webLinkFont $\backslash$URL \{url\}]
      \mbox{}\\
      e.g. \  \URL{http://www.eg.org/EG/DL/WS}

\item[\webLinkFont $\backslash$MailTo \{Email addr\}]
      \mbox{}\\
      e.g. \  \MailTo{publishing@eg.org}

\item[\webLinkFont $\backslash$MailToNA \{emailName\}\{@emailSiteAddress\}]
      \mbox{}\\
      e.g. \  \MailToNA{publishing}{@eg.org}

\item[\webLinkFont $\backslash$webLink\{URL without hyperlink creation\}]
      \mbox{}\\
      e.g. \  \webLink{http://www.eg.org/some_arbitrary_long/but_useless/URL}

\end{description}

%------------------------------------------------------------------------
\subsection{PDF Generation}

Your final paper should be delivered as a PDF document with all typefaces
embedded. \LaTeX{} users should use \texttt{dvips} and \texttt{ps2pdf} to
create this PDF document. Adobe Acrobat Distiller may be used in place of
\texttt{ps2pdf}.

Adobe PDFWriter is \emph{not} acceptable for use. Documents created with
PDFWriter will be returned to the author for revision. \texttt{pdftex} and
\texttt{pdflatex} (and its variants) can be used only if the author can
make certain that all typefaces are embedded and images are not downsampled
or subsampled during the PDF creation process.

Users with no access to these PDF creation tools should make available a
PostScript file and we will make a PDF document from it.

The PDF file \emph{must not} be change protected.

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: dvips / ps2pdf / etc.}

\noindent
\texttt{dvips} should be invoked with the \texttt{-Ppdf} and \texttt{-G0}
flags in order to use Type 1 PostScript typefaces:

\begin{verbatim}
    dvips -t a4 -Ppdf -G0 -o my.ps my.dvi
\end{verbatim}

\noindent
If you are using version 7.x of GhostScript, please use the following method of invoking \texttt{ps2pdf}, in
order to embed all typefaces and ensure that images are not downsampled or subsampled in the PDF
creation process:

\begin{verbatim}
  ps2pdf -dMaxSubsetPct=100 \
         -dCompatibilityLevel=1.3 \
         -dSubsetFonts=true \
         -dEmbedAllFonts=true \
         -dAutoFilterColorImages=false \
         -dAutoFilterGrayImages=false \
         -dColorImageFilter=/FlateEncode \
         -dGrayImageFilter=/FlateEncode \
         -dMonoImageFilter=/FlateEncode \
         mypaper.ps mypaper.pdf
\end{verbatim}

If you are using version 8.x of GhostScript, please use this method in place of the example above:
\begin{verbatim}
  ps2pdf -dPDFSETTINGS=/prepress \
         -dCompatibilityLevel=1.3 \
         -dAutoFilterColorImages=false \
         -dAutoFilterGrayImages=false \
         -dColorImageFilter=/FlateEncode \
         -dGrayImageFilter=/FlateEncode \
         -dMonoImageFilter=/FlateEncode \
         -dDownsampleColorImages=false \
         -dDownsampleGrayImages=false \
         mypaper.ps mypaper.pdf
\end{verbatim}

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: pdftex / pdflatex / etc.}

\noindent
Configuration of these tools to embed all typefaces can be accomplished by editing the \texttt{updmap.cfg} file
to enable inclusion of the standard (or base) 14 typefaces.

Linux users can run the \texttt{updmap} script to do this:
\begin{verbatim}
updmap --setoption pdftexDownloadBase14 true
\end{verbatim}

Windows users should edit the \texttt{updmap.cfg} files found in their TeX installation directories (one or both
of the following may be present):
\begin{verbatim}
  INSTALLDIR\texmf\web2c\updmap.cfg
  INSTALLDIR\localtexmf\miktex\config\updmap.cfg
\end{verbatim}

Ensure the value for \texttt{pdftexDownloadBase14} is "true," and then follow the instructions found here:
\httpAddr{//docs.miktex.org/manual/} to update your MikTeX installation.

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: Acrobat Distiller}

We recommend to use a Distiller job options file that embeds
all typefaces and does not downsample or subsample images when creating the PDF document.
%------------------------------------------------------------------------
\subsection{Exclusive License Form}

You must include your signed Eurographics Exclusive License Form 
when you submit your finished paper. We MUST have this form before
your paper can be published in the proceedings.

%-------------------------------------------------------------------------
\subsection{Conclusions}

Please direct any questions to the production editor in charge of
these proceedings.

%-------------------------------------------------------------------------

%\bibliographystyle{eg-alpha}
\bibliographystyle{eg-alpha-doi}

\bibliography{references}

%-------------------------------------------------------------------------
\newpage

\begin{figure*}[tbp]
  \centering
  \mbox{} \hfill
  % the following command controls the width of the embedded PS file
  % (relative to the width of the current column)
  \includegraphics[width=.3\linewidth]{sampleFig}
  % replacing the above command with the one below will explicitly set
  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
  % It will also prevent LaTeX from reading the PS file to determine
  % the bounding box (i.e., it will speed up the compilation process)
  % \includegraphics[width=.3\linewidth, bb=39 696 126 756]{sampleFig}
  \hfill
  \includegraphics[width=.3\linewidth]{sampleFig}
  \hfill \mbox{}
  \caption{\label{fig:ex3}%
           For publications with color tables (i.e., publications not offering
           color throughout the paper) please \textbf{observe}: 
           for the printed version -- and ONLY for the printed
           version -- color figures have to be placed in the last page.
           \newline
           For the electronic version, which will be converted to PDF before
           making it available electronically, the color images should be
           embedded within the document. Optionally, other multimedia
           material may be attached to the electronic version. }
\end{figure*}

\end{document}

### 2021-05-10 content/content_without_teaser.tex
init

--- 2021-05-10 content/content_with_teaser.tex
@@
 
-\title[Seminar Topic]%
-      {Seminar Topic about Graph Layouting for Visualizing Graph Structured Data}
+\title[EG \LaTeX\ Author Guidelines]%
+      {\LaTeX\ Author Guidelines for EUROGRAPHICS Proceedings Manuscripts}
 
@@
 % instead of the author's name (and leave the affiliation blank) !!
-\author[M.Rosier \& K.\"ohlein]
-{\parbox{\textwidth}{\centering Marcel Rosier$^1$% 
-\\ Seminar: Data Visualization %
-\\ marcel.rosier@tum.de% 
-\\ Supervisor: Kevin H\"ohlein%
+\author[D. Fellner \& S. Behnke]
+{\parbox{\textwidth}{\centering D.\,W. Fellner\thanks{Chairman Eurographics Publications Board}$^{1,2}$
+        and S. Behnke$^{2}$ 
 %        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
@@
 % For Computer Graphics Forum: Please use the abbreviation of your first name.
-{\parbox{\textwidth}{\centering $^1$ Technische Universit\"at M\"unchen
+{\parbox{\textwidth}{\centering $^1$TU Darmstadt \& Fraunhofer IGD, Germany\\
+         $^2$Graz University of Technology, Institute of Computer Graphics and Knowledge Visualization, Austria
 %        $^2$ Another Department to illustrate the use in papers from authors
@@
 
-\teaser{
- \includegraphics[width=\linewidth]{eg_new}
- \centering
-  \caption{New EG Logo}
-\label{fig:teaser}
-}
+% uncomment for using teaser
+% \teaser{
+%  \includegraphics[width=\linewidth]{eg_new}
+%  \centering
+%   \caption{New EG Logo}
+% \label{fig:teaser}
+%}
 
@@
 %-------------------------------------------------------------------------
+%  ACM CCS 2012
+   (see http://www.acm.org/about/class/class/2012)
+%The tool at \url{http://dl.acm.org/ccs.cfm} can be used to generate
+% CCS codes.
+%Example:
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10010147.10010371.10010352.10010381</concept_id>
+<concept_desc>Computing methodologies~Collision detection</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10010583.10010588.10010559</concept_id>
+<concept_desc>Hardware~Sensors and actuators</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10010583.10010584.10010587</concept_id>
+<concept_desc>Hardware~PCB design and layout</concept_desc>
+<concept_significance>100</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
 
+\ccsdesc[300]{Computing methodologies~Collision detection}
+\ccsdesc[300]{Hardware~Sensors and actuators}
+\ccsdesc[100]{Hardware~PCB design and layout}
+
+\printccsdesc   
 \end{abstract}  
@@
 end of your paper in alphabetical order. When referenced in the text, enclose
-the citation index in square brackets, for example~\cite{herman2000graph}. Where
+the citation index in square brackets, for example~\cite{Lous90}. Where
 appropriate, include the name(s) of editors of referenced books.
@@
 \item \textbf{one} author: first 3 chars plus year -- 
-      e.g.\ \cite{Herman2000} , \cite{VonLandesberger2011} \cite{Goyal2018} \cite{Lhuillier2017} \cite{Vehlow2015} \cite{Gibson2013}
+      e.g.\ \cite{Lous90}
 \item \textbf{two}, \textbf{three} or \textbf{four} authors: first char
@@
 
-\bibliography{references}
+\bibliography{egbibsample}
 

### 2021-05-10 content/paper.tex
init

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v2.02, Jan 25, 2017

\title[Seminar Topic]%
      {Graph Layouting for Visualizing Graph Structured Data}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
\author[M.Rosier \& K.\"ohlein]
{\parbox{\textwidth}{\centering Marcel Rosier$^1$% 
\\ Seminar: Data Visualization %
\\ marcel.rosier@tum.de% 
\\ Supervisor: Kevin H\"ohlein%
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
        }
        \\
% For Computer Graphics Forum: Please use the abbreviation of your first name.
{\parbox{\textwidth}{\centering $^1$ Technische Universit\"at M\"unchen
%        $^2$ Another Department to illustrate the use in papers from authors
%             with different affiliations
       }
}
}
% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{36}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page
\[\]

%-------------------------------------------------------------------------
\begin{document}

\teaser{
 \includegraphics[width=\linewidth]{media/example_teaser.pdf}
 \centering
  \caption{Edge Bundling teaser, currently just a placeholder, modified from \cite{Lhuillier2017}}
\label{fig:teaser}
}

\maketitle
%-------------------------------------------------------------------------
\begin{abstract}
   TODO

%-------------------------------------------------------------------------
%  ACM CCS 1998
%  (see http://www.acm.org/about/class/1998)
% \begin{classification} % according to http:http://www.acm.org/about/class/1998
% \CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
% \end{classification}
%-------------------------------------------------------------------------

\end{abstract}  
%-------------------------------------------------------------------------
\section{Introduction}
A graph is mathematically defined as a tuple $ G = (V, E)$ with $V$ being a set of vertices (often also referred to as nodes) and $E \subseteq V^2$ being a set of edges \cite{Al-Taie2017}.[Graphs are used for representation of data, bridge from pure math to data vis.] Each edge describes a relation between two nodes and can be either directed or undirected. A sequence of edges along distinct vertices is called a path. Moreover Edges and vertices can have attributes, that provide additional information about them. Graphs can also be time dependent and modifiable, e.g. through user interaction. These graphs are called dynamic. Due to the scope of this paper only static graphs will be considered. 
\\\\
% Graphs come in various forms: Reaching from simple hierarchical Trees, e.g. in a file system, to gigantic graphs with millions of nodes and edges, for instance a social network.
In various areas - such as Biology, Software Engineering or the analysis of Social Networks - graphs are a typical way of representing data. In order to analyze the stored information, it is necessary to visualize it. Therefore the mathematical concept of a graph needs to be mapped to a visual representation. This process isn't trivial since except for special cases like geographic data vertices have no fixed position.
The procedure of mapping a graph to a usually 2 or 3 dimensional area is called graph layouting. 
Considering the size of typical graphs that can quickly exceed thousands of nodes and edges - for example a graph representing a social network like Facebook -  the layout of a graph has a huge impact on understanding the local and global structure, connections and other analysis tasks. Thus layout techniques try to maximize readability by taking different aesthetic criteria into account. To further improve readability the graph can be reduced in its complexity. Typical methods are for instance filtering, grouping or edge bundling (e.g. \autoref{fig:teaser}) which can be applied algorithmically, by a user or combined. \\ This paper is organized as follows: Section 2 gives an overview of possible visual representations of a graph and different families of layouting algorithms. In Section 3 state of the art complexity reduction strategies are explained with a heavy focus on edge bundling techniques and section 4 outlines common user interaction methods. Finally Section 5 summarizes the main aspects and points to open research topics.

\section{Visual Representation}
\subsection{Visualization options and criteria}
Graph Visualizations usually come in two types: node-link diagrams and adjacency matrices.
Node-link diagrams map vertices to items and edges to links between the items. Positioning the nodes and drawing the links in an appropriate way, e.g. minimize edge crossings, is challenging. Adjacency matrices on the other side have no issues with edge crossings. In this approach a graph with $N$ vertices is visualized by a $N * N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column. To optimize readability an optimal node order has to be calculated, the edges are then implicitly displayed.
Both types have certain benefits: In a direct comparison node-link diagrams are more intuitive, compact and make it easier to follow paths but can be hard to understand due to problematic layouts. Adjacency matrices provide an uncluttered view for dense graphs but have issues displaying large networks \cite{Ghoniem2004}. Some approaches also combine the 2 types to synthesize their advantages. Examples for all mentioned options can be found in \autoref{fig:vis_types}.
% This representation provides an intuitive overview and allows users to easily follow paths \cite{}
\\
\\
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{media/vis_types.pdf}
    \caption{bad quality -> improve and adapt or find other figure; \cite{VonLandesberger2011}}
    \label{fig:vis_types}
\end{figure}

\begin{itemize}
    \item whats a good layout? aesthetic criteria 
\end{itemize}

\subsection{Layout Techniques}
\begin{itemize}
    \item algorithms for special graph types: e.g. trees 
    \item force based families
    \item constraint based
    \item multi-scale approaches
    \item Non standard layouts: e.g. dimension reduction
\end{itemize}

\section{Complexity Reduction}

\subsection{Preprocessing}
Filtering and aggregation
\subsection{Grouping/ Clustering}

\subsection{Edge Bundling}
Explain the core idea of edge bundling and different methods
\begin{itemize}
    \item Techniques/ Methods
    \item Visualizations: Blending, Data color mapping, Shading, Smoothing and Deformation
    \item Discussion: Pro/con, mention its possible that non existent patterns are recognized
\end{itemize}

\section{User Interaction}
\begin{itemize}
    \item Panning, Zooming
    \item Magic Lense, Fisheye, Focus+context
    \item Adapt visualization parameters
\end{itemize}

\section{Conclusions}
Citations so they show up in the references:
\cite{Herman2000}
\cite{VonLandesberger2011}
\cite{Goyal2018}
\cite{Lhuillier2017}
\cite{Vehlow2015}
\cite{Gibson2013} \\

%-------------------------------------------------------------------------

%\bibliographystyle{eg-alpha}
\bibliographystyle{eg-alpha-doi}

\bibliography{references}

%-------------------------------------------------------------------------

\end{document}

### 2021-05-27 content/paper.tex
- added first draft - added media files

@@
 % \pStartPage{1}      % set starting page
-\[\]
 
@@
  \centering
-  \caption{Edge Bundling teaser, currently just a placeholder, modified from \cite{Lhuillier2017}}
+  \caption{Global edge bundle relaxation . Taken from \cite{Lhuillier2017}}
 \label{fig:teaser}
@@
 \begin{abstract}
-   TODO
+Graph structured data can be found in biology, social networks and various other modern applications.
+To properly analyze this data and extract relevant information from it, a fitting visualization is necessary.
+Although basic graph drawing has been a research topic for a long time, the underlying graphs have grown to a scale, at which modern applications deal with huge graph structures that contain millions of nodes and edges.
+Therefore finding scalable approaches to produce comprehensible visualizations becomes more and more relevant.
+This paper provides an overview of basic graph representations, layouting and complexity reduction techniques as well as common user interaction methods.
+A special focus is given to the complexity reduction approach edge bundling.
 
@@
 \section{Introduction}
-A graph is mathematically defined as a tuple $ G = (V, E)$ with $V$ being a set of vertices (often also referred to as nodes) and $E \subseteq V^2$ being a set of edges \cite{Al-Taie2017}.[Graphs are used for representation of data, bridge from pure math to data vis.] Each edge describes a relation between two nodes and can be either directed or undirected.
+A graph is mathematically defined as a tuple $ G = (V, E)$ with $V$ being a set of vertices (often also referred to as nodes) and $E \subseteq V^2$ being a set of edges \cite{Al-Taie2017}.In the context of using graphs for data representation each edge describes a relation between two nodes and can be either directed or undirected.
 A sequence of edges along distinct vertices is called a path.
 Moreover Edges and vertices can have attributes, that provide additional information about them.
+A special type of graph that is connected and contains no cycles is called a tree.
 Graphs can also be time dependent and modifiable, e.g.
@@
 These graphs are called dynamic.
-Due to the scope of this paper only static graphs will be considered.
-\\\\
+Due to the scope of this paper only static graphs will be considered.\\
+\\
 % Graphs come in various forms: Reaching from simple hierarchical Trees, e.g. in a file system, to gigantic graphs with millions of nodes and edges, for instance a social network.
 In various areas - such as Biology, Software Engineering or the analysis of Social Networks - graphs are a typical way of representing data.
-In order to analyze the stored information, it is necessary to visualize it.
+In order to analyze the contained information, it is necessary to visualize it.
 Therefore the mathematical concept of a graph needs to be mapped to a visual representation.
-This process isn't trivial since except for special cases like geographic data vertices have no fixed position.
+This process isn't trivial since for most applications the vertices have no fixed positions.
 The procedure of mapping a graph to a usually 2 or 3 dimensional area is called graph layouting. 
@@
 Finally Section 5 summarizes the main aspects and points to open research topics.
-
 \section{Visual Representation}
@@
 In this approach a graph with $N$ vertices is visualized by a $N * N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
-To optimize readability an optimal node order has to be calculated, the edges are then implicitly displayed.
+To optimize readability an optimal node order has to be calculated, the edges are then implicitly displayed.\\
 Both types have certain benefits: In a direct comparison node-link diagrams are more intuitive, compact and make it easier to follow paths but can be hard to understand due to problematic layouts.
-Adjacency matrices provide an uncluttered view for dense graphs but have issues displaying large networks \cite{Ghoniem2004}.
+Adjacency matrices contrarily provide an uncluttered view for dense graphs but have issues displaying large networks \cite{Ghoniem2004}.
 Some approaches also combine the 2 types to synthesize their advantages.
 Examples for all mentioned options can be found in \autoref{fig:vis_types}.
+\\\\
+In most cases node-link diagrams are used.
+This visualization style has infinitely many representations for the same graph that differ in readability.
+Therefore criteria that define a good layout are necessary to later on guide layouting techniques to good results.
+While many exist that each benefit the visualization of certain graphs, a few guidelines are common and work well on most graphs: \\
+Nodes should be evenly distributed and vertices that are similar with respect to attributes or edges should be clustered.
+Uniform edge lengths, minimal edge crossings and bends as well as an overall symmetric layout are also favorable \cite{Bennett2007}.
+\\
+While all of these goals contribute to a good layout, its often impossible to achieve all of them simultaneously since they are competing with each other\cite{Gibson2013}.
+Hence a few criteria have to be prioritized and as a result different layout techniques focus on different goals.
+%or even ignore them altogether.
+In the following subsection a selection of layout algorithms classified by their node placement will be introduced and shortly reviewed.
 % This representation provides an intuitive overview and allows users to easily follow paths \cite{}
-\\
-\\
 \begin{figure}
@@
     \includegraphics[width=\linewidth]{media/vis_types.pdf}
-    \caption{bad quality -> improve and adapt or find other figure; \cite{VonLandesberger2011}}
+    \caption{Three common types of graph visualization techniques. Taken from \cite{VonLandesberger2011}}
     \label{fig:vis_types}
@@
 
-\begin{itemize}
-    \item whats a good layout? aesthetic criteria 
-\end{itemize}
+\subsection{Layout Techniques}
+One of the oldest and most established classes is the family of force-based layouts. In this class spring-electrical and energy based approaches have to be distinguished.\\ 
+The first type makes use of Eades' spring-embedder, meaning that nodes are modeled as steel rings and edges as springs.
+Connected nodes attract each other while all other nodes repel.
+After initializing the system with a random state, the attracting and repulsive forces and the systems temperature are repeatedly calculated to update node positions until the system reaches an equilibrium.
+Although this approach was only intended to work for small graphs (< 50 nodes), many adaptions have been made that tried to improve both run-time and the resulting layout.
+Among other adaptions, for instance, Fruchterman and Reingold sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration.
+\\
+Energy based approaches also utilize Eades' spring-embedder but consider the layout task as an optimisation problem in which an energy function that contains the desired properties of the layout has to be minimized.
+For Example Kamada and Kawai try to approximate the layouts' euclidean distance with their graph-theoretic distance (the shortest path length between 2 vertices).
+Symmetry is the main goal of their approach and consequently their cost function measure the lack of it.
+\\
+Despite the various adaptions and run-time improvements this class of algorithms still isn't suited for large and dense graphs.\cite{Gibson2013} \cite{VonLandesberger2011}.
+\\
+Constraint-based layouts extend the idea of force based approaches.
+They allow the user to define constraints such as closeness of groups, node alignment along an axis, edge direction etc.
+While this extremely enhances the power of expression it also exacerbates the run-time issue of the underlying force-based approach\cite{VonLandesberger2011}.\\
+Another category are multiscale algortihms.
+The core idea is to hierarchically decompose a graph into smaller sub-graphs and iteratively add nodes to an initially rough structure.
+These algorithms differ in their graph decomposition and layer layouting technique.
+For instance Muelder and Ma used clustering for decomposition and space filling curves for the layout, which resulted in short run-times and comprehensible layouts even for dense graphs  \cite{Muelder2008}.
+Typically this class of algorithms is significantly faster than force-based techniques\cite{VonLandesberger2011}.
+\\
+Besides these rather established techniques also non standard algorithms such as dimension reduction approaches can be used to generate layouts.
+This type of algorithm tries to map information that's expressed in a high-dimensional space onto a mostly 2 or 3 dimensional space.
+The preservation of relevant information has to be guaranteed as far as possible during this reduction.
+For this cause the graph-theoretic distance between node pairs is often used as the information to be saved.
+While layouting graphs fast, the produced layouts' quality highly depends on the graphs structure and is quite sensitive, meaning that small structural modifications significantly influence the layouts quality\cite{Gibson2013} \cite{VonLandesberger2011}.
 
-\subsection{Layout Techniques}
-\begin{itemize}
-    \item algorithms for special graph types: e.g. trees 
-    \item force based families
-    \item constraint based
-    \item multi-scale approaches
-    \item Non standard layouts: e.g. dimension reduction
-\end{itemize}
+\begin{figure}
+    \centering
+    \includegraphics{media/space_filling_trees.pdf}
+    \caption{Examples for space filling tree visualizations while using the principle of a) enclosure and b) adjacency. Taken from \cite{VonLandesberger2011}}
+    \label{fig:space_filling_trees}
+\end{figure}
+
+All previously introduced techniques work on general graphs.
+However, in certain use-cases graphs have a specific structure that allow specialized techniques to produces high quality layouts with reduced complexity.
+For instance file systems or taxonomies are typically represented as a tree.
+Trees can be visualized with node-link diagrams exemplary in 2D using layered, radial or balloon layouts, in 3D with cone trees or with hyperbolic layouts.
+Herman et al.
+provides a good overview of the mentioned techniques \cite{Herman2000}.
+Most of them have a linear complexity in time and memory which makes them extremely scalable and hence preferable to general layout techniques if applicable.
+\\
+In addition to the node-link visualizations especially trees can also be displayed with space-filling techniques.
+These try to make use of the entire space and utilize e.g.
+the principle of enclosure (recursively layout children within the area of the parent) or adjacency (place children next to parents without overlapping) to display edges instead of links\cite{VonLandesberger2011}.
+Examples for space filling trees can be found in \autoref{fig:space_filling_trees}\\
+No matter how good a layout is - if a graph contains thousands or even millions of nodes and edges it gets extremely hard to understand its structure and extract valuable information.
+For this purpose different complexity reduction techniques can be applied that either highlight relevant information or remove details for the sake of an better overview.
+A selection of such techniques will be presented in the following section.
 
@@
 
-\subsection{Preprocessing}
-Filtering and aggregation
-\subsection{Grouping/ Clustering}
+\subsection{Pre-processing}
+Before using a layout technique on a graph pre-processing algorithms are often applied to simplify the graph while preserving its main structure.
+These simplifications can usually be performed autonomous without user interaction and come in 2 main forms:\\
+Graph filtering and graph aggregation.
+Graph Filtering can be either stochastic, which means that nodes and edges are mainly removed randomly, or deterministic.
+Deterministic algorithms also remove nodes and edges but make that decisions based on node or edge attributes, topological values or other graph properties.\\
+Graph aggregation on the other hand does not remove nodes and edges but merges them and thereby reduces size and complexity.
+This aggregation step can be applied multiple times to generate different levels of detail of the same graph.
+The selection process of nodes that will be aggregated is usually based on attributes, clusters or known hierarchies \cite{VonLandesberger2011}.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.5]{media/group_vis.pdf}
+    \caption{Different grouping techniques for the same graph.
+(a) Visual node attributes - here color.
+(b) Juxtaposed - here using an attached approach.
+(c) Superimposed - here using a contour approach.
+(d) Embedded - here using a hybrid approach.
+Taken from \cite{Vehlow2015}}
+    \label{fig:group_vis}
+\end{figure}
+\subsection{Grouping}
+Another highly useful way of reducing complexity is to highlight group structures to provide a better overview of the data.
+For this purpose four main visualization types exist: Visual node attributes, Juxtaposed, Superimposed and Embedded.\\
+The first technique, called 'visual node attributes', uses the nodes visual representation properties to encode its association with one or multiple groups.
+Regardless of the fact that humans can't easily distinguish more than 7 colors \cite{Healey1996}, node coloring is often used to link vertices to groups.
+Different techniques try to optimize the color selection by either maximizing color differences between neighbouring groups or assigning similar colors to similar groups.
+Visualizing groups via visual node attributes is commonly combined with other approaches that will be introduced in the following.\\
+Techniques that display the graph and its group structure adjacently are referred to as juxtaposed approaches.
+In this category one has to further differentiate between attached and separate visualizations.
+Attached visualizations directly connect to the graph representation and are usually applied to adjacency matrix visualizations.
+Even tough they can be applied to node-link diagrams, if the nodes are aligned along one axis, this condition makes the approach unsuitable for large node-link representations.
+Separate juxtaposed visualizations are more appropriate for node-link diagrams.
+They visualize the group structure in a separate view and then connect the both representations via visual indicators, like e.g.
+color or links, or interaction.\\
+Superimposed approaches render an additional layer on top of the graph visualization that is aligned with the underlying node-link diagram.
+To illustrate the groups one of three categories can be used: Line overlay methods draw a continuous line between all nodes of each group.
+In order to reduce cluttering, the lines can, for example, be minimized by applying an adapted travelling sales man heuristic \cite{Alper2011}.
+Contour overlays on the other hand draw closed contours that can differ in shape.
+All nodes within a contour are considered to be part of the group that is represented by the respective surrounding shape.
+The last type, Partitioning, is contrarily to the previous technique a space filling approach that decomposes the area into sections.
+Each of these section behaves identical to the contour idea by associating nodes with groups through visual enclosing.
+All introduced superimposed techniques are usually combined with coloring.\\
+Lastly groups can be visualized through embedding.
+This means that groups are also modelled as nodes and are a part of the graph.
+Beside the standard node-link diagram, a hybrid node-link and matrix representation can be useful for this technique \cite{Vehlow2015}.\\
+Examples for all 4 introduced group visualization types can be found in \autoref{fig:group_vis}.
 
 \subsection{Edge Bundling}
-Explain the core idea of edge bundling and different methods
-\begin{itemize}
-    \item Techniques/ Methods
-    \item Visualizations: Blending, Data color mapping, Shading, Smoothing and Deformation
-    \item Discussion: Pro/con, mention its possible that non existent patterns are recognized
-\end{itemize}
+The last here presented complexity reduction technique is called Edge Bundling.
+The main idea is to spatially group similar graph edges to reduce clutter and provide a better overview.
+Techniques for 3D graphs and also special graphs such as compound graphs exist.
+However, the following section will focus only on general two dimensional graphs and provide an overview of common techniques and challenges.
 
+\subsubsection{Bundling Techniques}
+Basis of most bundling techniques is a so-called compatibility function, typically named $\kappa$, that measures how similar paths are, meaning that low values of $\kappa$ correlate to similar paths and vice versa.
+Defining a good compatibility function that both considers the layouts spatial and the graphs attribute information is challenging.
+Moreover edge bundling techniques have to be differentiated between directed an undirected graphs\cite{Lhuillier2017}.
+
+\paragraph*{Undirected Graphs} Undirected graphs are the most general type for graph bundling algorithms.
+One of the first approaches was force-directed edge bundling (FDEB).
+It defines $\kappa$ only with respect to geometric information of the edge drawing.
+This basic technique was soon extended by Ngyuen et al.
+to also consider topology as well as attributes and therefore importance-based compatibility\cite{Nguyen2012}.
+While this modification grants more control about the bundling selection it still has no explicit control mesh.
+A control mesh is a set of control points that determines the rough structure of the resulting edges and thereby enables even more control about the resulting graph representation.
+A progression based on control meshes proposed by Luo et al.
+can help to deal with common bundling issues like difficult end-to-end edge tracing or closeness of bundled edges to unrelated nodes.
+They achieve this by adapting $\kappa$ to be zero for all edges that do not both share a node and are close in the base drawing.
+Moreover edges are repelled from unrelated nodes\cite{Luo2012}.
+While this extension eases edge tracing for small graphs (< 100 nodes and edges) it often fails to sufficiently reduce clutter for larger graphs with thousands of nodes and edges.\\
+All previously presented techniques have issues with large graphs either due to run-time (FDBE, \cite{Nguyen2012}) or insufficient clutter reduction \cite{Luo2012}.
+MINGLE \cite{Gansner2011} tackles these issues by introducing a scalable ink-saving principle that tries to minimize the theoretically used ink to draw edges by using a greedy algorithm that merges close edges as long as it saves ink.
+This allows bundling for graphs with up to a million edges with the drawback of creating less smooth results \cite{Lhuillier2017}.
+
+\paragraph*{Directed Graphs}  A common criticism of general bundling techniques is that they ignore the semantic carried by edge directions even though this information can be vital for certain applications such as exploring a software system's call graph.
+The additional information of direction can be easily added by coloring edges with a source to destination color gradient.
+However, this leads to color mixing problems when edges of different directions are bundled.
+To fix this problem, techniques specifically for directed graphs have been introduced that add the edge direction to the compatibility function $\kappa$.
+An example algorithm for directed edge bundling is Divided Edge Bundling (DEB)\cite{Selassie2011}, which can be seen as an extension of FDEB that takes edge direction into account.
+Among other modifications it adds repulsion of opposing edges and includes edge weight into $\kappa$ to reduce bundling of important edges.
+This method produces good results for directed graphs but claims more space and produces more clutter compared to FDEB.
+A direct comparison between the both approaches can be found in \autoref{fig:airline_edge_bundling} \cite{Lhuillier2017}.\\
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=1]{media/airline_edge_bundling.pdf}
+    \caption{Comparison of FDEB and DEB for the US airlines graph ($|V| = 235, |E| = 2101$). Taken from\cite{Lhuillier2017}}
+    \label{fig:airline_edge_bundling}
+\end{figure}
+
+\subsubsection{Bundling Visualizations}
+After applying a bundling technique the bundled data can be visualized and even further improved using different techniques such as blending, data color mapping, shading or smoothing and deformation.\\
+Blending deals with overlapping edges and eases the readability of connection strength or amount of paths between endpoints.
+This is achieved by rendering dense bundles more opaque than sparse ones by mapping the local bundle density to visual properties like opacity, color or shading (\autoref{fig:blending}) \cite{Lhuillier2017}.\\
+Data color mapping maps edge attributes to colors.
+As already seen in section 3.3.1 - directed graphs, this attribute could be direction but moreover also spatial density or domain specific attributes like exemplary the local flight height for the us airline example.
+As mentioned earlier overlapping paths provoke a color mix issue for which no general solution exists so far.
+To further emphasize important connections shading can be used.
+For instance a false height map that is increasing parabolically from a bundles border to its center can be defined.
+This map is then processed by a shading algorithm, making crossing edge easier to distinguish.
+This is often combined with color mapping.
+Finally bundled edges can also be smoothed, commonly achieved by applying Laplacian smoothing, or deformed in order to avoid landmarks in the drawing space.
+The latter can be accomplished e.g.
+by making the landmarks repulsive, which can both be applied during the bundling iterations or as a post-processing step \cite{Lhuillier2017}.
+
+\begin{figure}
+    \centering
+    \includegraphics{media/blending.pdf}
+    \caption{A graph drawn with opacity blending (right) and without (left). Taken from \cite{Lhuillier2017}}
+    \label{fig:blending}
+\end{figure}
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.8]{media/faithfullness_bundling.pdf}
+    \caption{Pseudo-random graph (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}}
+    \label{fig:faithfulness}
+\end{figure}
+\subsubsection{Bundling Challenges}
+Additional to the already discussed difficulties of bundling, a few other general challenges like quality, faithfulness and control have to be mentioned.
+In general it's quite hard to define what qualifies a bundling as good.
+Although certain criteria such as ink-minimization, edge crossings, faithfulness and more have been discussed, no such set of metrics entirely covers edge bundling\cite{Lhuillier2017}.
+Moreover edge bundling techniques have to be evaluated on how much information is lost during the process and also if wrong information is added, e.g.
+non-existent structures (\autoref{fig:faithfulness}).
+Lastly, due to advances in the field, bundling techniques nowadays have a lot of parameters that are often coupled.
+This aggravates the recreation of graphs for users after switching techniques.\\
+Despite multiple general complexity reduction techniques, resulting graph layouts often still don't fit a users specific need. 
+For this purpose a range of user interaction techniques exists that enables the user to further enhance the visualization and interact with it.
+A selection of such approaches will be presented in the following.
+
+\begin{figure}
+    \centering
+    \includegraphics{media/edge_lense.pdf}
+    \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens From. Taken from \cite{VonLandesberger2011}}
+    \label{fig:edge_lense}
+\end{figure}
 \section{User Interaction}
-\begin{itemize}
-    \item Panning, Zooming
-    \item Magic Lense, Fisheye, Focus+context
-    \item Adapt visualization parameters
-\end{itemize}
+The most traditional tools of user interaction are zoom and pan.
+Zooming can take two forms: Geometric zooming simply blows up the graph content while semantic zooming also changes the content by adding details\cite{Herman2000}.
+To generate different level of detail for the latter, aggregation techniques as described in section 3.1 can be used.
+Panning on the other hand allows the user to translate the plot to different view points in a zoomed in graph.
+For node-link diagrams panning is often implemented to move along the graphs edges (guided panning) to properly explore the graphs structure \cite{VonLandesberger2011}.\\
+While zooming can help to focus on a specific part of the data, contextual information is lost in the process, which makes it hard to fully understand the displayed information.
+To solve this issue a technique often referred to as magic-lense or focus+context can be used.
+The main idea is to highlight an area of interest by assigning more space to the data within it and distorting the surrounding.
+This idea can be applied to node-link diagrams as well as space-filling techniques and also allows the user to select multiple areas of focus.
+For geometric changes this technique is called fisheye-views.
+These so-called lenses can be either applied to the entire graph data within the selected area or just the edges (edge lenses).
+Edge lenses help to declutter strongly overlapping edges in graphs that try to minimize node repositioning (e.g.
+geographic-based graphs as in \autoref{fig:edge_lense}).
+Other approaches use edge-filtering in the focused area or move neighbouring nodes closer to a highlighted vertex.\\
+In Addition to modifying the final result of a graph visualization, users can also adjust the underlying visualization type and its parameters.
+Typical changes to visual parameters are highlighting of interesting items, which is often synchronized across multiple views of the same graph, or changes to the visual schemes layout or representation.
+Layout adaptions can be either changes to the layout type that are followed by an automatic recalculation or manual movement of nodes.
+User can also select a different visual representation, e.g.
+switching from a node-link diagram to a matrix representation, which can provide new insights.
+If applicable, smooth animations should be used to make the transformations comprehensible \cite{VonLandesberger2011}.
 
 \section{Conclusions}
-Citations so they show up in the references:
-\cite{Herman2000}
-\cite{VonLandesberger2011}
-\cite{Goyal2018}
-\cite{Lhuillier2017}
-\cite{Vehlow2015}
-\cite{Gibson2013} \\
+Graph layouting is a challenging topic that made significant progress within the last decades.
+It is used in demanded areas such as biology, social network analysis and more and heavily contributes to extracting useful information from huge sets of graph structured data.
+This paper collated information from prominent research and provided a brief overview of various graph representation types and introduced different categories of layouting techniques.
+Moreover multiple approaches to reduce a graphs complexity, namely pre-processing steps like filtering and aggregation, grouping methods and especially edge bundling techniques and their visualizations and challenges have been presented.
+Lastly different user interaction methods have been discussed.
+Although recent techniques have become more scalable, this process needs to continue to enable the analysis of the data sets that keep growing.
+However, approaches still need to remain usable in practice by  limiting their usage complexity to a reasonable level.
+%layouting and complexity reduction techniques as well as different user interaction possibilities by collating information from prominent research.
 

### 2021-05-27 content/paper.tex
- implemented frist feedback, added figure

@@
 \teaser{
- \includegraphics[width=\linewidth]{media/example_teaser.pdf}
+ \includegraphics[width=\linewidth]{media/teaser.pdf}
  \centering
-  \caption{Global edge bundle relaxation . Taken from \cite{Lhuillier2017}}
+  \caption{Four frames from a global edge bundle relaxation process. The bundling tightness increases from a to d. Taken from \cite{Lhuillier2017}.  }
 \label{fig:teaser}
@@
 Graph structured data can be found in biology, social networks and various other modern applications.
-To properly analyze this data and extract relevant information from it, a fitting visualization is necessary.
+To get an overview of this data and extract relevant information from it, a fitting visualization is necessary.
 Although basic graph drawing has been a research topic for a long time, the underlying graphs have grown to a scale, at which modern applications deal with huge graph structures that contain millions of nodes and edges.
@@
 \section{Introduction}
-A graph is mathematically defined as a tuple $ G = (V, E)$ with $V$ being a set of vertices (often also referred to as nodes) and $E \subseteq V^2$ being a set of edges \cite{Al-Taie2017}.In the context of using graphs for data representation each edge describes a relation between two nodes and can be either directed or undirected.
+A graph is mathematically defined as a tuple $ G = (V, E)$ with $V$ being a set of vertices (often also referred to as nodes) and $E \subseteq V^2$ being a set of edges \cite{Al-Taie2017}.
+In the context of using graphs for data representation each edge describes a relation between two nodes and can be either directed or undirected.
 A sequence of edges along distinct vertices is called a path.
-Moreover Edges and vertices can have attributes, that provide additional information about them.
-A special type of graph that is connected and contains no cycles is called a tree.
+Moreover edges and vertices can have attributes, that provide additional information about them.
+A connected acyclic graph is called a tree.
 Graphs can also be time dependent and modifiable, e.g.
 through user interaction.
-These graphs are called dynamic.
-Due to the scope of this paper only static graphs will be considered.\\
+These graphs are called dynamic.Visualizing dynamic graphs, however, is a topic on its own and beyond the scope of this paper.\\
 \\
@@
 In various areas - such as Biology, Software Engineering or the analysis of Social Networks - graphs are a typical way of representing data.
-In order to analyze the contained information, it is necessary to visualize it.
+In order to analyze the contained information, it is insightful to visualize it.
 Therefore the mathematical concept of a graph needs to be mapped to a visual representation.
-This process isn't trivial since for most applications the vertices have no fixed positions.
+This process isn't trivial since because in many applications there may not be a natural way of mapping graphs with potentially high-dimensional and abstract node and edge properties to 2- or 3-dimensional visual representations.
 The procedure of mapping a graph to a usually 2 or 3 dimensional area is called graph layouting. 
@@
 Thus layout techniques try to maximize readability by taking different aesthetic criteria into account.
-To further improve readability the graph can be reduced in its complexity.
+To improve readability the graph can be reduced in its complexity.
 Typical methods are for instance filtering, grouping or edge bundling (e.g.
-\autoref{fig:teaser}) which can be applied algorithmically, by a user or combined.
+\autoref{fig:teaser}) which can be applied algorithmically, manually through user interactions or combined.
 \\ This paper is organized as follows: Section 2 gives an overview of possible visual representations of a graph and different families of layouting algorithms.
-In Section 3 state of the art complexity reduction strategies are explained with a heavy focus on edge bundling techniques and section 4 outlines common user interaction methods.
+In Section 3 state of the art complexity reduction strategies are explained with a focus on edge bundling techniques and section 4 outlines common user interaction methods.
 Finally Section 5 summarizes the main aspects and points to open research topics.
+
 \section{Visual Representation}
 \subsection{Visualization options and criteria}
-Graph Visualizations usually come in two types: node-link diagrams and adjacency matrices.
-Node-link diagrams map vertices to items and edges to links between the items.
+This paper will focus on two types of graph visualizations: node-link diagrams and adjacency matrices.
+Node-link diagrams map nodes to positions and edges to links between the node positions.
 Positioning the nodes and drawing the links in an appropriate way, e.g.
 minimize edge crossings, is challenging.
-Adjacency matrices on the other side have no issues with edge crossings.
-In this approach a graph with $N$ vertices is visualized by a $N * N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
-To optimize readability an optimal node order has to be calculated, the edges are then implicitly displayed.\\
+Adjacency matrices on the other side don't have issues with edge crossings.
+In this approach a graph with $N$ vertices is visualized by a $N \times N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
+To improve readability an optimized node order has to be calculated, the edges are then implicitly displayed.\\
 Both types have certain benefits: In a direct comparison node-link diagrams are more intuitive, compact and make it easier to follow paths but can be hard to understand due to problematic layouts.
 Adjacency matrices contrarily provide an uncluttered view for dense graphs but have issues displaying large networks \cite{Ghoniem2004}.
-Some approaches also combine the 2 types to synthesize their advantages.
+Some approaches also combine the two types to synthesize their advantages.
 Examples for all mentioned options can be found in \autoref{fig:vis_types}.
 \\\\
-In most cases node-link diagrams are used.
+Most of the techniques introduced in this paper use node-link diagrams.
 This visualization style has infinitely many representations for the same graph that differ in readability.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.16]{media/layout_techniques.pdf}
+    \caption{The protein interaction network laid out using the graph embedder (GEM) (a) and Kamada and Kawai's algorithm (b). Modified from \cite{Gibson2013}}
+    \label{fig:layout_techniques}
+\end{figure}
 \subsection{Layout Techniques}
 One of the oldest and most established classes is the family of force-based layouts. In this class spring-electrical and energy based approaches have to be distinguished.\\ 
-The first type makes use of Eades' spring-embedder, meaning that nodes are modeled as steel rings and edges as springs.
+The first type makes use of Eades' spring-embedder \cite{Gibson2013}, meaning that nodes are modeled as steel rings and edges as springs.
 Connected nodes attract each other while all other nodes repel.
@@
 Although this approach was only intended to work for small graphs (< 50 nodes), many adaptions have been made that tried to improve both run-time and the resulting layout.
-Among other adaptions, for instance, Fruchterman and Reingold sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration.
-\\
-Energy based approaches also utilize Eades' spring-embedder but consider the layout task as an optimisation problem in which an energy function that contains the desired properties of the layout has to be minimized.
-For Example Kamada and Kawai try to approximate the layouts' euclidean distance with their graph-theoretic distance (the shortest path length between 2 vertices).
+Among other adaptions, for instance, Fruchterman and Reingold\cite{not accesible!} sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration.\\
+Energy based approaches also utilize Eades' spring-embedder but consider the layout task as an optimisation problem in which an energy function that contains the desired properties of the layout has to be minimized \cite{Gibson2013}.
+For Example Kamada and Kawai \cite{Kamada1989} try to approximate the layouts' Euclidean distance with their graph-theoretic distance (the shortest path length between two vertices).
 Symmetry is the main goal of their approach and consequently their cost function measure the lack of it.
-\\
+Visual examples for the mentioned algorithms can be found in \autoref{fig:layout_techniques}.\\
 Despite the various adaptions and run-time improvements this class of algorithms still isn't suited for large and dense graphs.\cite{Gibson2013} \cite{VonLandesberger2011}.
@@
 They allow the user to define constraints such as closeness of groups, node alignment along an axis, edge direction etc.
-While this extremely enhances the power of expression it also exacerbates the run-time issue of the underlying force-based approach\cite{VonLandesberger2011}.\\
+This enhances the power of expression but exacerbates the run-time issue of the underlying force-based approach\cite{VonLandesberger2011}.
+
 Another category are multiscale algortihms.
@@
 These algorithms differ in their graph decomposition and layer layouting technique.
-For instance Muelder and Ma used clustering for decomposition and space filling curves for the layout, which resulted in short run-times and comprehensible layouts even for dense graphs  \cite{Muelder2008}.
+For instance Muelder and Ma \cite{Muelder2008} used clustering for decomposition and space filling curves for the layout, which resulted in short run-times and comprehensible layouts even for dense graphs.
 Typically this class of algorithms is significantly faster than force-based techniques\cite{VonLandesberger2011}.
@@
 the principle of enclosure (recursively layout children within the area of the parent) or adjacency (place children next to parents without overlapping) to display edges instead of links\cite{VonLandesberger2011}.
-Examples for space filling trees can be found in \autoref{fig:space_filling_trees}\\
+Examples for space filling trees can be found in \autoref{fig:space_filling_trees}.\\
+
+\section{Complexity Reduction}
 No matter how good a layout is - if a graph contains thousands or even millions of nodes and edges it gets extremely hard to understand its structure and extract valuable information.
 For this purpose different complexity reduction techniques can be applied that either highlight relevant information or remove details for the sake of an better overview.
-A selection of such techniques will be presented in the following section.
-
-\section{Complexity Reduction}
+A selection of such techniques will be presented in this following section.
 
@@
 Before using a layout technique on a graph pre-processing algorithms are often applied to simplify the graph while preserving its main structure.
-These simplifications can usually be performed autonomous without user interaction and come in 2 main forms:\\
+These simplifications can usually be performed autonomously without user interaction and come in two main forms:\\
 Graph filtering and graph aggregation.
-Graph Filtering can be either stochastic, which means that nodes and edges are mainly removed randomly, or deterministic.
+Graph filtering can be either stochastically, which means that nodes and edges are mainly removed randomly, or deterministically.
 Deterministic algorithms also remove nodes and edges but make that decisions based on node or edge attributes, topological values or other graph properties.\\
@@
 Another highly useful way of reducing complexity is to highlight group structures to provide a better overview of the data.
-For this purpose four main visualization types exist: Visual node attributes, Juxtaposed, Superimposed and Embedded.\\
+For this purpose Vehlow et al.\cite{Vehlow2015} propose four main visualization types: Visual node attributes, Juxtaposed, Superimposed and Embedded.\\
 The first technique, called 'visual node attributes', uses the nodes visual representation properties to encode its association with one or multiple groups.
@@
 
-\paragraph*{Undirected Graphs} Undirected graphs are the most general type for graph bundling algorithms.
+\paragraph*{Undirected Graphs} Undirected graphs are a common type for graph bundling algorithms.
 One of the first approaches was force-directed edge bundling (FDEB).
@@
 The additional information of direction can be easily added by coloring edges with a source to destination color gradient.
-However, this leads to color mixing problems when edges of different directions are bundled.
+However, this leads to color mixing problems when edges of different directions are bundled \cite{Lhuillier2017}.
 To fix this problem, techniques specifically for directed graphs have been introduced that add the edge direction to the compatibility function $\kappa$.
@@
 This method produces good results for directed graphs but claims more space and produces more clutter compared to FDEB.
-A direct comparison between the both approaches can be found in \autoref{fig:airline_edge_bundling} \cite{Lhuillier2017}.\\
+A direct comparison between both approaches can be found in \autoref{fig:airline_edge_bundling} \cite{Lhuillier2017}.\\
 
@@
     \centering
-    \includegraphics[scale=1]{media/airline_edge_bundling.pdf}
+    \includegraphics[scale=0.9]{media/airline_edge_bundling.pdf}
     \caption{Comparison of FDEB and DEB for the US airlines graph ($|V| = 235, |E| = 2101$). Taken from\cite{Lhuillier2017}}
@@
 Data color mapping maps edge attributes to colors.
-As already seen in section 3.3.1 - directed graphs, this attribute could be direction but moreover also spatial density or domain specific attributes like exemplary the local flight height for the us airline example.
+As already seen in section 3.3.1 - directed graphs, this attribute could be direction but moreover also spatial density or domain specific attributes like for example the local flight height for the us airline example.
 As mentioned earlier overlapping paths provoke a color mix issue for which no general solution exists so far.
@@
 \subsubsection{Bundling Challenges}
-Additional to the already discussed difficulties of bundling, a few other general challenges like quality, faithfulness and control have to be mentioned.
+In addition to the already discussed difficulties of bundling, a few other general challenges like quality, faithfulness and control have to be mentioned.
 In general it's quite hard to define what qualifies a bundling as good.
@@
 This idea can be applied to node-link diagrams as well as space-filling techniques and also allows the user to select multiple areas of focus.
-For geometric changes this technique is called fisheye-views.
+For geometric changes this technique is called fisheye-views \cite{VonLandesberger2011}.
 These so-called lenses can be either applied to the entire graph data within the selected area or just the edges (edge lenses).
@@
 Other approaches use edge-filtering in the focused area or move neighbouring nodes closer to a highlighted vertex.\\
-In Addition to modifying the final result of a graph visualization, users can also adjust the underlying visualization type and its parameters.
+In addition to modifying the final result of a graph visualization, users can also adjust the underlying visualization type and its parameters.
 Typical changes to visual parameters are highlighting of interesting items, which is often synchronized across multiple views of the same graph, or changes to the visual schemes layout or representation.
@@
 Graph layouting is a challenging topic that made significant progress within the last decades.
-It is used in demanded areas such as biology, social network analysis and more and heavily contributes to extracting useful information from huge sets of graph structured data.
-This paper collated information from prominent research and provided a brief overview of various graph representation types and introduced different categories of layouting techniques.
+It is used in areas such as biology, social network analysis and more and heavily contributes to extracting useful information from huge sets of graph structured data.
+This paper collated information from various research and provided a brief overview of various graph representation types and introduced different categories of layouting techniques.
 Moreover multiple approaches to reduce a graphs complexity, namely pre-processing steps like filtering and aggregation, grouping methods and especially edge bundling techniques and their visualizations and challenges have been presented.

### 2021-06-06 content/paper.tex
src check, figure descr improvment, proof read

@@
 Graph structured data can be found in biology, social networks and various other modern applications.
-To get an overview of this data and extract relevant information from it, a fitting visualization is necessary.
+To get an overview of this data and extract relevant information from it a fitting visualization is necessary.
 Although basic graph drawing has been a research topic for a long time, the underlying graphs have grown to a scale, at which modern applications deal with huge graph structures that contain millions of nodes and edges.
-Therefore finding scalable approaches to produce comprehensible visualizations becomes more and more relevant.
+Therefore finding appropriate layouts and reasonable simplification techniques for the sake of a better overview becomes more and more relevant.
 This paper provides an overview of basic graph representations, layouting and complexity reduction techniques as well as common user interaction methods.
@@
 Therefore the mathematical concept of a graph needs to be mapped to a visual representation.
-This process isn't trivial since because in many applications there may not be a natural way of mapping graphs with potentially high-dimensional and abstract node and edge properties to 2- or 3-dimensional visual representations.
+This process isn't trivial since in many applications there may not be a natural way of mapping graphs with potentially high-dimensional and abstract node and edge properties to 2- or 3-dimensional visual representations.
 The procedure of mapping a graph to a usually 2 or 3 dimensional area is called graph layouting. 
@@
 In this approach a graph with $N$ vertices is visualized by a $N \times N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
-To improve readability an optimized node order has to be calculated, the edges are then implicitly displayed.\\
+To improve readability an optimized node order has to be calculated, the edges are then implicitly displayed \cite{VonLandesberger2011}.\\
 Both types have certain benefits: In a direct comparison node-link diagrams are more intuitive, compact and make it easier to follow paths but can be hard to understand due to problematic layouts.
@@
 While many exist that each benefit the visualization of certain graphs, a few guidelines are common and work well on most graphs: \\
-Nodes should be evenly distributed and vertices that are similar with respect to attributes or edges should be clustered.
+For instance nodes should be evenly distributed and vertices that are similar with respect to attributes or edges should be clustered.
 Uniform edge lengths, minimal edge crossings and bends as well as an overall symmetric layout are also favorable \cite{Bennett2007}.
@@
     \includegraphics[width=\linewidth]{media/vis_types.pdf}
-    \caption{Three common types of graph visualization techniques. Taken from \cite{VonLandesberger2011}}
+    \caption{Three common types of graph visualization techniques (node-link, adjacency matrix and a combination) applied to the same graph. Taken from \cite{VonLandesberger2011}}
     \label{fig:vis_types}
@@
 Although this approach was only intended to work for small graphs (< 50 nodes), many adaptions have been made that tried to improve both run-time and the resulting layout.
-Among other adaptions, for instance, Fruchterman and Reingold\cite{not accesible!} sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration.\\
+Among other adaptions, for instance, Fruchterman and Reingold\cite{Fruchterman1991} sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration \cite{Gibson2013}.\\
 Energy based approaches also utilize Eades' spring-embedder but consider the layout task as an optimisation problem in which an energy function that contains the desired properties of the layout has to be minimized \cite{Gibson2013}.
 For Example Kamada and Kawai \cite{Kamada1989} try to approximate the layouts' Euclidean distance with their graph-theoretic distance (the shortest path length between two vertices).
-Symmetry is the main goal of their approach and consequently their cost function measure the lack of it.
+Symmetry is the main goal of their approach and consequently their cost function measures the lack of it.
 Visual examples for the mentioned algorithms can be found in \autoref{fig:layout_techniques}.\\
-Despite the various adaptions and run-time improvements this class of algorithms still isn't suited for large and dense graphs.\cite{Gibson2013} \cite{VonLandesberger2011}.
+Despite the various adaptions and run-time improvements, this class of algorithms still isn't suited for large and dense graphs.\cite{Gibson2013} \cite{VonLandesberger2011}.
 \\
@@
 These try to make use of the entire space and utilize e.g.
-the principle of enclosure (recursively layout children within the area of the parent) or adjacency (place children next to parents without overlapping) to display edges instead of links\cite{VonLandesberger2011}.
-Examples for space filling trees can be found in \autoref{fig:space_filling_trees}.\\
+the principle of enclosure (recursively layout children within the area of the parent, \autoref{fig:space_filling_trees}a) or adjacency (place children next to parents without overlapping, \autoref{fig:space_filling_trees}b) to display edges instead of links\cite{VonLandesberger2011}.
 
@@
 \subsection{Pre-processing}
-Before using a layout technique on a graph pre-processing algorithms are often applied to simplify the graph while preserving its main structure.
-These simplifications can usually be performed autonomously without user interaction and come in two main forms:\\
-Graph filtering and graph aggregation.
+Before using a layout technique on a graph, pre-processing algorithms are often applied to simplify the graph while preserving its core structure.
+These simplifications can usually be performed autonomously without user interaction and come in two main forms: Graph filtering and graph aggregation \cite{VonLandesberger2011}.
 Graph filtering can be either stochastically, which means that nodes and edges are mainly removed randomly, or deterministically.
@@
 For this purpose Vehlow et al.\cite{Vehlow2015} propose four main visualization types: Visual node attributes, Juxtaposed, Superimposed and Embedded.\\
-The first technique, called 'visual node attributes', uses the nodes visual representation properties to encode its association with one or multiple groups.
+The first technique, called 'visual node attributes', uses the nodes visual representation properties to encode its association with one or multiple groups ( \autoref{fig:group_vis}a).
 Regardless of the fact that humans can't easily distinguish more than 7 colors \cite{Healey1996}, node coloring is often used to link vertices to groups.
@@
 Visualizing groups via visual node attributes is commonly combined with other approaches that will be introduced in the following.\\
-Techniques that display the graph and its group structure adjacently are referred to as juxtaposed approaches.
+Techniques that display the graph and its group structure adjacently are referred to as juxtaposed approaches (\autoref{fig:group_vis}b).
 In this category one has to further differentiate between attached and separate visualizations.
@@
 Each of these section behaves identical to the contour idea by associating nodes with groups through visual enclosing.
-All introduced superimposed techniques are usually combined with coloring.\\
+All introduced superimposed techniques are usually combined with coloring (\autoref{fig:group_vis}c).\\
 Lastly groups can be visualized through embedding.
-This means that groups are also modelled as nodes and are a part of the graph.
+This means that groups are also modelled as nodes and are a part of the graph (\autoref{fig:group_vis}d).
 Beside the standard node-link diagram, a hybrid node-link and matrix representation can be useful for this technique \cite{Vehlow2015}.\\
-Examples for all 4 introduced group visualization types can be found in \autoref{fig:group_vis}.
 
 \subsection{Edge Bundling}
-The last here presented complexity reduction technique is called Edge Bundling.
+The last here presented complexity reduction technique is called edge bundling.
 The main idea is to spatially group similar graph edges to reduce clutter and provide a better overview.
@@
 \subsubsection{Bundling Techniques}
-Basis of most bundling techniques is a so-called compatibility function, typically named $\kappa$, that measures how similar paths are, meaning that low values of $\kappa$ correlate to similar paths and vice versa.
+Basis of most bundling techniques is a so-called compatibility function, typically named $\kappa$ that measures how similar paths are, meaning that low values of $\kappa$ correlate to similar paths and vice versa.
 Defining a good compatibility function that both considers the layouts spatial and the graphs attribute information is challenging.
@@
 This basic technique was soon extended by Ngyuen et al.
-to also consider topology as well as attributes and therefore importance-based compatibility\cite{Nguyen2012}.
+\cite{Nguyen2011} to also consider topology as well as attributes and therefore importance-based compatibility.
 While this modification grants more control about the bundling selection it still has no explicit control mesh.
@@
 A progression based on control meshes proposed by Luo et al.
-can help to deal with common bundling issues like difficult end-to-end edge tracing or closeness of bundled edges to unrelated nodes.
+\cite{Luo2012} can help to deal with common bundling issues like difficult end-to-end edge tracing or closeness of bundled edges to unrelated nodes.
 They achieve this by adapting $\kappa$ to be zero for all edges that do not both share a node and are close in the base drawing.
-Moreover edges are repelled from unrelated nodes\cite{Luo2012}.
+Moreover edges are repelled from unrelated nodes.
 While this extension eases edge tracing for small graphs (< 100 nodes and edges) it often fails to sufficiently reduce clutter for larger graphs with thousands of nodes and edges.\\
-All previously presented techniques have issues with large graphs either due to run-time (FDBE, \cite{Nguyen2012}) or insufficient clutter reduction \cite{Luo2012}.
-MINGLE \cite{Gansner2011} tackles these issues by introducing a scalable ink-saving principle that tries to minimize the theoretically used ink to draw edges by using a greedy algorithm that merges close edges as long as it saves ink.
+All previously presented techniques have issues with large graphs either due to run-time (FDBE, \cite{Nguyen2011}) or insufficient clutter reduction (\cite{Luo2012}).
+MINGLE \cite{Gansner2011} tackles these issues by introducing a computationally-scalable ink-saving principle that tries to minimize the theoretically used ink to draw edges by using a greedy algorithm that merges close edges as long as it saves ink.
 This allows bundling for graphs with up to a million edges with the drawback of creating less smooth results \cite{Lhuillier2017}.
@@
     \includegraphics[scale=0.9]{media/airline_edge_bundling.pdf}
-    \caption{Comparison of FDEB and DEB for the US airlines graph ($|V| = 235, |E| = 2101$). Taken from\cite{Lhuillier2017}}
+    \caption{Comparison of a general bundling algorithm (FDEB, a) and a bundling algorithm for directed graphs (DEB, b) applied on the US airlines graph ($|V| = 235, |E| = 2101$).
+Taken from\cite{Lhuillier2017}}
     \label{fig:airline_edge_bundling}
@@
 After applying a bundling technique the bundled data can be visualized and even further improved using different techniques such as blending, data color mapping, shading or smoothing and deformation.\\
-Blending deals with overlapping edges and eases the readability of connection strength or amount of paths between endpoints.
+Blending deals with overlapping edges and eases the readability of connection strength and amount of paths between endpoints.
 This is achieved by rendering dense bundles more opaque than sparse ones by mapping the local bundle density to visual properties like opacity, color or shading (\autoref{fig:blending}) \cite{Lhuillier2017}.\\
@@
 For instance a false height map that is increasing parabolically from a bundles border to its center can be defined.
-This map is then processed by a shading algorithm, making crossing edge easier to distinguish.
+This map is then processed by a shading algorithm, making crossing edges easier to distinguish.
 This is often combined with color mapping.
@@
     \centering
-    \includegraphics{media/blending.pdf}
+    \includegraphics[scale=0.9]{media/blending.pdf}
     \caption{A graph drawn with opacity blending (right) and without (left). Taken from \cite{Lhuillier2017}}
@@
     \centering
-    \includegraphics[scale=0.8]{media/faithfullness_bundling.pdf}
-    \caption{Pseudo-random graph (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}}
+    \includegraphics[scale=0.7]{media/faithfullness_bundling.pdf}
+    \caption{A pseudo-random graph visualized without simplification (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}}
     \label{fig:faithfulness}
@@
 Although certain criteria such as ink-minimization, edge crossings, faithfulness and more have been discussed, no such set of metrics entirely covers edge bundling\cite{Lhuillier2017}.
-Moreover edge bundling techniques have to be evaluated on how much information is lost during the process and also if wrong information is added, e.g.
-non-existent structures (\autoref{fig:faithfulness}).
+Moreover edge bundling techniques have to be evaluated on how much information is lost during the process and also if wrong information is added.
+For instance bundling techniques can visualize non-existent structures in a pseudo-random graph (\autoref{fig:faithfulness}).
 Lastly, due to advances in the field, bundling techniques nowadays have a lot of parameters that are often coupled.
-This aggravates the recreation of graphs for users after switching techniques.\\
+This aggravates the recreation of graphs for users after switching techniques \cite{Lhuillier2017}.\\
 Despite multiple general complexity reduction techniques, resulting graph layouts often still don't fit a users specific need. 
@@
     \centering
-    \includegraphics{media/edge_lense.pdf}
-    \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens From. Taken from \cite{VonLandesberger2011}}
+    \includegraphics[scale=0.8]{media/edge_lense.pdf}
+    \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens. Taken from \cite{VonLandesberger2011}}
     \label{fig:edge_lense}
@@
 While zooming can help to focus on a specific part of the data, contextual information is lost in the process, which makes it hard to fully understand the displayed information.
-To solve this issue a technique often referred to as magic-lense or focus+context can be used.
+To solve this issue a technique often referred to as magic-lense \cite{VonLandesberger2011} or focus+context \cite{Herman2000} can be used.
 The main idea is to highlight an area of interest by assigning more space to the data within it and distorting the surrounding.
 This idea can be applied to node-link diagrams as well as space-filling techniques and also allows the user to select multiple areas of focus.
-For geometric changes this technique is called fisheye-views \cite{VonLandesberger2011}.
+For geometric changes this technique is called fisheye views \cite{VonLandesberger2011}.
 These so-called lenses can be either applied to the entire graph data within the selected area or just the edges (edge lenses).
@@
 Graph layouting is a challenging topic that made significant progress within the last decades.
-It is used in areas such as biology, social network analysis and more and heavily contributes to extracting useful information from huge sets of graph structured data.
-This paper collated information from various research and provided a brief overview of various graph representation types and introduced different categories of layouting techniques.
+It is used in areas such as biology, social network analysis and more and contributes to extracting useful information from huge sets of graph structured data.
+This paper collated information from various research and provided a brief overview of graph representation types and introduced different categories of layouting techniques.
 Moreover multiple approaches to reduce a graphs complexity, namely pre-processing steps like filtering and aggregation, grouping methods and especially edge bundling techniques and their visualizations and challenges have been presented.

### 2021-06-09 content/paper.tex
src check, proof read, grammar check

@@
 Although basic graph drawing has been a research topic for a long time, the underlying graphs have grown to a scale, at which modern applications deal with huge graph structures that contain millions of nodes and edges.
-Therefore finding appropriate layouts and reasonable simplification techniques for the sake of a better overview becomes more and more relevant.
+Therefore finding appropriate layouts and reasonable simplification techniques becomes more and more relevant.
 This paper provides an overview of basic graph representations, layouting and complexity reduction techniques as well as common user interaction methods.
@@
 A sequence of edges along distinct vertices is called a path.
-Moreover edges and vertices can have attributes, that provide additional information about them.
+Moreover, edges and vertices can have attributes, that provide additional information about them.
 A connected acyclic graph is called a tree.
@@
 through user interaction.
-These graphs are called dynamic.Visualizing dynamic graphs, however, is a topic on its own and beyond the scope of this paper.\\
+These graphs are called dynamic.
+Visualizing dynamic graphs, however, is a topic on its own and beyond the scope of this paper.\\
 \\
 % Graphs come in various forms: Reaching from simple hierarchical Trees, e.g. in a file system, to gigantic graphs with millions of nodes and edges, for instance a social network.
-In various areas - such as Biology, Software Engineering or the analysis of Social Networks - graphs are a typical way of representing data.
+In various areas - such as biology, software engineering or the analysis of social networks - graphs are a typical way of representing data.
 In order to analyze the contained information, it is insightful to visualize it.
@@
 This process isn't trivial since in many applications there may not be a natural way of mapping graphs with potentially high-dimensional and abstract node and edge properties to 2- or 3-dimensional visual representations.
-The procedure of mapping a graph to a usually 2 or 3 dimensional area is called graph layouting. 
-Considering the size of typical graphs that can quickly exceed thousands of nodes and edges - for example a graph representing a social network like Facebook -  the layout of a graph has a huge impact on understanding the local and global structure, connections and other analysis tasks.
+The procedure of mapping a graph to a usually 2 or 3-dimensional area is called graph layouting. 
+Considering the size of typical graphs that can quickly exceed thousands of nodes and edges - for example, a graph representing a social network like facebook -  the layout of a graph has a huge impact on understanding the local and global structure, connections and other analysis tasks.
 Thus layout techniques try to maximize readability by taking different aesthetic criteria into account.
@@
 \\ This paper is organized as follows: Section 2 gives an overview of possible visual representations of a graph and different families of layouting algorithms.
-In Section 3 state of the art complexity reduction strategies are explained with a focus on edge bundling techniques and section 4 outlines common user interaction methods.
-Finally Section 5 summarizes the main aspects and points to open research topics.
+In section 3 state of the art complexity reduction strategies are explained with a focus on edge bundling techniques and section 4 outlines common user interaction methods.
+Finally, section 5 summarizes the main aspects and points to open research topics.
 
@@
 Adjacency matrices on the other side don't have issues with edge crossings.
-In this approach a graph with $N$ vertices is visualized by a $N \times N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
+In this approach, a graph with $N$ vertices is visualized by an $N \times N $ matrix, in which every entry represents the edge between the vertices of the corresponding row and column.
 To improve readability an optimized node order has to be calculated, the edges are then implicitly displayed \cite{VonLandesberger2011}.\\
@@
 This visualization style has infinitely many representations for the same graph that differ in readability.
-Therefore criteria that define a good layout are necessary to later on guide layouting techniques to good results.
+Therefore criteria that define a good layout are necessary to, later on, guide layouting techniques to good results.
 While many exist that each benefit the visualization of certain graphs, a few guidelines are common and work well on most graphs: \\
-For instance nodes should be evenly distributed and vertices that are similar with respect to attributes or edges should be clustered.
+For instance, nodes should be evenly distributed and vertices that are similar with respect to attributes or edges should be clustered.
 Uniform edge lengths, minimal edge crossings and bends as well as an overall symmetric layout are also favorable \cite{Bennett2007}.
@@
 %or even ignore them altogether.
-In the following subsection a selection of layout algorithms classified by their node placement will be introduced and shortly reviewed.
+In the following subsection, a selection of layout algorithms classified by their node placement will be introduced and shortly reviewed.
 % This representation provides an intuitive overview and allows users to easily follow paths \cite{}
@@
     \includegraphics[width=\linewidth]{media/vis_types.pdf}
-    \caption{Three common types of graph visualization techniques (node-link, adjacency matrix and a combination) applied to the same graph. Taken from \cite{VonLandesberger2011}}
+    \caption{Three common types of graph visualization techniques (node-link, adjacency matrix and a combination) applied to the same graph. Taken from \cite{VonLandesberger2011}.}
     \label{fig:vis_types}
@@
     \includegraphics[scale=0.16]{media/layout_techniques.pdf}
-    \caption{The protein interaction network laid out using the graph embedder (GEM) (a) and Kamada and Kawai's algorithm (b). Modified from \cite{Gibson2013}}
+    \caption{The protein interaction network laid out using the graph embedder (GEM) (a) and Kamada and Kawai's algorithm (b). Modified from \cite{Gibson2013}.}
     \label{fig:layout_techniques}
@@
 Connected nodes attract each other while all other nodes repel.
-After initializing the system with a random state, the attracting and repulsive forces and the systems temperature are repeatedly calculated to update node positions until the system reaches an equilibrium.
+After initializing the system with a random state, the attracting and repulsive forces and the system's temperature are repeatedly calculated to update node positions until the system reaches an equilibrium.
 Although this approach was only intended to work for small graphs (< 50 nodes), many adaptions have been made that tried to improve both run-time and the resulting layout.
@@
 
-Another category are multiscale algortihms.
+Another category is multiscale algorithms.
 The core idea is to hierarchically decompose a graph into smaller sub-graphs and iteratively add nodes to an initially rough structure.
@@
 The preservation of relevant information has to be guaranteed as far as possible during this reduction.
-For this cause the graph-theoretic distance between node pairs is often used as the information to be saved.
-While layouting graphs fast, the produced layouts' quality highly depends on the graphs structure and is quite sensitive, meaning that small structural modifications significantly influence the layouts quality\cite{Gibson2013} \cite{VonLandesberger2011}.
+For this cause, the graph-theoretic distance between node pairs is often used as the information to be saved.
+While layouting graphs fast, the produced layouts' quality highly depends on the graph's structure and is quite sensitive, meaning that small structural modifications significantly influence the layout's quality \cite{Gibson2013} \cite{VonLandesberger2011}.
 
@@
     \includegraphics{media/space_filling_trees.pdf}
-    \caption{Examples for space filling tree visualizations while using the principle of a) enclosure and b) adjacency. Taken from \cite{VonLandesberger2011}}
+    \caption{Examples for space filling tree visualizations while using the principle of a) enclosure and b) adjacency. Taken from \cite{VonLandesberger2011}.}
     \label{fig:space_filling_trees}
@@
 All previously introduced techniques work on general graphs.
-However, in certain use-cases graphs have a specific structure that allow specialized techniques to produces high quality layouts with reduced complexity.
-For instance file systems or taxonomies are typically represented as a tree.
+However, in certain use-cases graphs have a specific structure that allow specialized techniques to produce high quality layouts with reduced complexity.
+For instance, file systems or taxonomies are typically represented as a tree.
 Trees can be visualized with node-link diagrams exemplary in 2D using layered, radial or balloon layouts, in 3D with cone trees or with hyperbolic layouts.
 Herman et al.
-provides a good overview of the mentioned techniques \cite{Herman2000}.
+provides an overview of the mentioned techniques \cite{Herman2000}.
 Most of them have a linear complexity in time and memory which makes them extremely scalable and hence preferable to general layout techniques if applicable.
@@
 Graph filtering can be either stochastically, which means that nodes and edges are mainly removed randomly, or deterministically.
-Deterministic algorithms also remove nodes and edges but make that decisions based on node or edge attributes, topological values or other graph properties.\\
+Deterministic algorithms also remove nodes and edges but make these decisions based on node or edge attributes, topological values or other graph properties.\\
 Graph aggregation on the other hand does not remove nodes and edges but merges them and thereby reduces size and complexity.
@@
 (d) Embedded - here using a hybrid approach.
-Taken from \cite{Vehlow2015}}
+Taken from \cite{Vehlow2015}.}
     \label{fig:group_vis}
@@
 \subsection{Grouping}
-Another highly useful way of reducing complexity is to highlight group structures to provide a better overview of the data.
+Another highly useful way of reducing complexity is to highlight group structures to provide an better overview of the data.
 For this purpose Vehlow et al.\cite{Vehlow2015} propose four main visualization types: Visual node attributes, Juxtaposed, Superimposed and Embedded.\\
@@
 Techniques that display the graph and its group structure adjacently are referred to as juxtaposed approaches (\autoref{fig:group_vis}b).
-In this category one has to further differentiate between attached and separate visualizations.
+In this category, one has to further differentiate between attached and separate visualizations.
 Attached visualizations directly connect to the graph representation and are usually applied to adjacency matrix visualizations.
@@
 The last type, Partitioning, is contrarily to the previous technique a space filling approach that decomposes the area into sections.
-Each of these section behaves identical to the contour idea by associating nodes with groups through visual enclosing.
-All introduced superimposed techniques are usually combined with coloring (\autoref{fig:group_vis}c).\\
+Each of these section behaves identically to the contour idea by associating nodes with groups through visual enclosing.
+All introduced superimposed techniques can be combined with coloring (\autoref{fig:group_vis}c).\\
 Lastly groups can be visualized through embedding.
@@
 The last here presented complexity reduction technique is called edge bundling.
-The main idea is to spatially group similar graph edges to reduce clutter and provide a better overview.
+The main idea is to spatially group similar graph edges to reduce clutter and provide an better overview.
 Techniques for 3D graphs and also special graphs such as compound graphs exist.
@@
 \subsubsection{Bundling Techniques}
-Basis of most bundling techniques is a so-called compatibility function, typically named $\kappa$ that measures how similar paths are, meaning that low values of $\kappa$ correlate to similar paths and vice versa.
+Basis of most bundling techniques is a so-called compatibility function, typically named $\kappa$ that measures how dissimilar paths are, meaning that low values of $\kappa$ correlate to similar paths and vice versa.
 Defining a good compatibility function that both considers the layouts spatial and the graphs attribute information is challenging.
-Moreover edge bundling techniques have to be differentiated between directed an undirected graphs\cite{Lhuillier2017}.
+Moreover, edge bundling techniques have to be differentiated between directed an undirected graphs\cite{Lhuillier2017}.
 
@@
 \cite{Luo2012} can help to deal with common bundling issues like difficult end-to-end edge tracing or closeness of bundled edges to unrelated nodes.
-They achieve this by adapting $\kappa$ to be zero for all edges that do not both share a node and are close in the base drawing.
-Moreover edges are repelled from unrelated nodes.
+They achieve this by adapting $\kappa$ to be non-null only for edges that share a node and are close in the base drawing.
+Moreover, edges are repelled from unrelated nodes.
 While this extension eases edge tracing for small graphs (< 100 nodes and edges) it often fails to sufficiently reduce clutter for larger graphs with thousands of nodes and edges.\\
@@
 An example algorithm for directed edge bundling is Divided Edge Bundling (DEB)\cite{Selassie2011}, which can be seen as an extension of FDEB that takes edge direction into account.
-Among other modifications it adds repulsion of opposing edges and includes edge weight into $\kappa$ to reduce bundling of important edges.
+Among other modifications, it adds repulsion of opposing edges and includes edge weight into $\kappa$ to reduce bundling of important edges.
 This method produces good results for directed graphs but claims more space and produces more clutter compared to FDEB.
@@
     \caption{Comparison of a general bundling algorithm (FDEB, a) and a bundling algorithm for directed graphs (DEB, b) applied on the US airlines graph ($|V| = 235, |E| = 2101$).
-Taken from\cite{Lhuillier2017}}
+Taken from\cite{Lhuillier2017}.}
     \label{fig:airline_edge_bundling}
@@
 As mentioned earlier overlapping paths provoke a color mix issue for which no general solution exists so far.
-To further emphasize important connections shading can be used.
-For instance a false height map that is increasing parabolically from a bundles border to its center can be defined.
+To further emphasize important connections, shading can be used.
+For instance, a false height map that is increasing parabolically from a bundles border to its center can be defined.
 This map is then processed by a shading algorithm, making crossing edges easier to distinguish.
@@
     \includegraphics[scale=0.9]{media/blending.pdf}
-    \caption{A graph drawn with opacity blending (right) and without (left). Taken from \cite{Lhuillier2017}}
+    \caption{A graph drawn with opacity blending (right) and without (left). Taken from \cite{Lhuillier2017}.}
     \label{fig:blending}
@@
     \includegraphics[scale=0.7]{media/faithfullness_bundling.pdf}
-    \caption{A pseudo-random graph visualized without simplification (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}}
+    \caption{A pseudo-random graph visualized without simplification (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}.}
     \label{fig:faithfulness}
@@
 In addition to the already discussed difficulties of bundling, a few other general challenges like quality, faithfulness and control have to be mentioned.
-In general it's quite hard to define what qualifies a bundling as good.
+In general, it is quite hard to define what qualifies a bundling as good.
 Although certain criteria such as ink-minimization, edge crossings, faithfulness and more have been discussed, no such set of metrics entirely covers edge bundling\cite{Lhuillier2017}.
-Moreover edge bundling techniques have to be evaluated on how much information is lost during the process and also if wrong information is added.
-For instance bundling techniques can visualize non-existent structures in a pseudo-random graph (\autoref{fig:faithfulness}).
+Moreover, edge bundling techniques have to be evaluated on how much information is lost during the process and also if wrong information is added.
+For instance, bundling techniques can visualize non-existent structures in a pseudo-random graph (\autoref{fig:faithfulness}).
 Lastly, due to advances in the field, bundling techniques nowadays have a lot of parameters that are often coupled.
 This aggravates the recreation of graphs for users after switching techniques \cite{Lhuillier2017}.\\
-Despite multiple general complexity reduction techniques, resulting graph layouts often still don't fit a users specific need. 
-For this purpose a range of user interaction techniques exists that enables the user to further enhance the visualization and interact with it.
+Despite multiple general complexity reduction techniques, resulting graph layouts often still don't fit a user's specific need. 
+For this purpose, a range of user interaction techniques exists that enable the user to further enhance the visualization and interact with it.
 A selection of such approaches will be presented in the following.
@@
     \includegraphics[scale=0.8]{media/edge_lense.pdf}
-    \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens. Taken from \cite{VonLandesberger2011}}
+    \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens. Taken from \cite{VonLandesberger2011}.}
     \label{fig:edge_lense}
@@
 To generate different level of detail for the latter, aggregation techniques as described in section 3.1 can be used.
-Panning on the other hand allows the user to translate the plot to different view points in a zoomed in graph.
-For node-link diagrams panning is often implemented to move along the graphs edges (guided panning) to properly explore the graphs structure \cite{VonLandesberger2011}.\\
+Panning on the other hand allows the user to translate the plot to different viewpoints in a zoomed in graph.
+For node-link diagrams panning is often implemented to move along the graph's edges (guided panning) to properly explore the graph's structure \cite{VonLandesberger2011}.\\
 While zooming can help to focus on a specific part of the data, contextual information is lost in the process, which makes it hard to fully understand the displayed information.
@@
 Moreover multiple approaches to reduce a graphs complexity, namely pre-processing steps like filtering and aggregation, grouping methods and especially edge bundling techniques and their visualizations and challenges have been presented.
-Lastly different user interaction methods have been discussed.
+Lastly, different user interaction methods have been discussed.
 Although recent techniques have become more scalable, this process needs to continue to enable the analysis of the data sets that keep growing.

### 2021-06-13 content/paper.tex
added missing src, adapted layout to fit 6 pages

@@
     \centering
-    \includegraphics[scale=0.16]{media/layout_techniques.pdf}
+    \includegraphics[scale=0.14]{media/layout_techniques.pdf}
     \caption{The protein interaction network laid out using the graph embedder (GEM) (a) and Kamada and Kawai's algorithm (b). Modified from \cite{Gibson2013}.}
@@
 Although this approach was only intended to work for small graphs (< 50 nodes), many adaptions have been made that tried to improve both run-time and the resulting layout.
-Among other adaptions, for instance, Fruchterman and Reingold\cite{Fruchterman1991} sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration \cite{Gibson2013}.\\
+Among other adaptions, for instance, Fruchterman and Reingold\cite{Fruchterman1991} sped up their algorithm by only considering neighbouring nodes within a certain radius for the force calculation and the graph embedder algorithm (GEM) \cite{Frick1995} additionally considered gravitational forces and moreover introduced a local temperature that determines how far a node can travel in a single iteration \cite{Gibson2013}.\\
 Energy based approaches also utilize Eades' spring-embedder but consider the layout task as an optimisation problem in which an energy function that contains the desired properties of the layout has to be minimized \cite{Gibson2013}.
@@
     \centering
-    \includegraphics{media/space_filling_trees.pdf}
+    \includegraphics[scale=1]{media/space_filling_trees.pdf}
     \caption{Examples for space filling tree visualizations while using the principle of a) enclosure and b) adjacency. Taken from \cite{VonLandesberger2011}.}
@@
     \centering
-    \includegraphics[scale=0.9]{media/airline_edge_bundling.pdf}
+    \includegraphics[scale=0.88]{media/airline_edge_bundling.pdf}
     \caption{Comparison of a general bundling algorithm (FDEB, a) and a bundling algorithm for directed graphs (DEB, b) applied on the US airlines graph ($|V| = 235, |E| = 2101$).
@@
     \centering
-    \includegraphics[scale=0.9]{media/blending.pdf}
+    \includegraphics[scale=0.82]{media/blending.pdf}
     \caption{A graph drawn with opacity blending (right) and without (left). Taken from \cite{Lhuillier2017}.}
@@
     \centering
-    \includegraphics[scale=0.7]{media/faithfullness_bundling.pdf}
+    \includegraphics[scale=0.65]{media/faithfullness_bundling.pdf}
     \caption{A pseudo-random graph visualized without simplification (a) and an exemplary bundling (b). Taken from \cite{Lhuillier2017}.}
@@
     \centering
-    \includegraphics[scale=0.8]{media/edge_lense.pdf}
+    \includegraphics[scale=0.7]{media/edge_lense.pdf}
     \caption{Example of edge lens interaction. (a) Original view without lens. (b) Using edge lens. Taken from \cite{VonLandesberger2011}.}
@@
 Panning on the other hand allows the user to translate the plot to different viewpoints in a zoomed in graph.
-For node-link diagrams panning is often implemented to move along the graph's edges (guided panning) to properly explore the graph's structure \cite{VonLandesberger2011}.\\
+For node-link diagrams panning is often implemented to move along the graph's edges (guided panning) to properly explore the graph's structure \cite{VonLandesberger2011}.
 While zooming can help to focus on a specific part of the data, contextual information is lost in the process, which makes it hard to fully understand the displayed information.
